---
title: "Practical 3"
output: html_notebook
---

```{r}
#read the package and layer
library(sf)
library(here)
st_layers(here("prac3_data", "gadm36_AUS.gpkg"))

Ausoutline <- st_read(here("prac3_data", "gadm36_AUS.gpkg"), 
                      layer='gadm36_AUS_0')

print(Ausoutline)
st_crs(Ausoutline)$proj4string

#only useful if there is no CRS
Ausoutline <- Ausoutline %>%
  st_set_crs(., 4326)
```

```{r}
# re-project or transform CRS
AusoutlinePROJECTED <- Ausoutline %>%
  st_transform(.,3112) # GDA94, a local CRS for Australia

print(AusoutlinePROJECTED)
```

```{r}
##From sf to sp
#AusoutlineSP <- Ausoutline %>%
#  as(., "Spatial")

##From sp to sf
#AusoutlineSF <- AusoutlineSP %>%
#  st_as_sf()
```

```{r}
library(raster)
library(terra)
jan<-terra::rast(here("prac3_data", "wc2.1_5m_tavg_01.tif"))
# have a look at the raster layer jan
jan
```

```{r}
plot(jan)
```

using the Mollweide projection saved to a new object. The Mollweide projection retains area proportions whilst compromising accuracy of angle and shape
```{r}
# set the proj 4 to a new object

pr1 <- terra::project(jan, "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs")

#or....

newproj<-"ESRI:54009"
# get the jan raster and give it the new proj4
pr1 <- jan %>%
  terra::project(., newproj)
plot(pr1)
```
```{r}
#back to WGS 84
pr1 <- pr1 %>%
  terra::project(., "EPSG:4326")
plot(pr1)
```

```{r}
# look in our folder, find the files that end with .tif and 
library(fs)
dir_info("D:/CASA/GIS/prac3_data/") 
```

```{r}
# select the data we actually want
library(tidyverse)
listfiles<-dir_info("D:/CASA/GIS/prac3_data/") %>%
  filter(str_detect(path, ".tif")) %>%
  dplyr::select(path)%>%
  pull()

#have a look at the file names 
listfiles
```

Then load all of the data straight into a SpatRaster. A SpatRaster is a collection of raster layers with the same spatial extent and resolution.
```{r}
worldclimtemp <- listfiles %>%
  terra::rast()
  
#have a look at the raster stack
worldclimtemp
```

```{r}
# access the january layer
worldclimtemp[[1]]
```
```{r}
#rename the layers
month <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
           "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

names(worldclimtemp) <- month
```

```{r}
# get data for January using new layer name
worldclimtemp$Jan
```

Using a raster stack we can extract data with a single command!! 
For example, make a dataframe of some sample sites — Australian cities/towns.
```{r}
site <- c("Brisbane", "Melbourne", "Perth", "Sydney", "Broome", "Darwin", "Orange", 
          "Bunbury", "Cairns", "Adelaide", "Gold Coast", "Canberra", "Newcastle", 
          "Wollongong", "Logan City" )
lon <- c(153.03, 144.96, 115.86, 151.21, 122.23, 130.84, 149.10, 115.64, 145.77, 
         138.6, 153.43, 149.13, 151.78, 150.89, 153.12)
lat <- c(-27.47, -37.91, -31.95, -33.87, 17.96, -12.46, -33.28, -33.33, -16.92, 
         -34.93, -28, -35.28, -32.93, -34.42, -27.64)
#Put all of this inforamtion into one list 
samples <- data.frame(site, lon, lat, row.names="site")
# Extract the data from the Rasterstack for all points 
AUcitytemp<- terra::extract(worldclimtemp, samples)
```

Add the city names to the rows of AUcitytemp
```{r}
Aucitytemp2 <- AUcitytemp %>% 
  as_tibble()%>% 
  add_column(Site = site, .before = "Jan")
```

take Perth as an example. We can subset our data either using the row name:
```{r}
Perthtemp <- Aucitytemp2 %>%
  filter(site=="Perth")
```

Make a histogram of Perth’s temperature. The tibble stored the data as double and the base hist() function needs it as numeric..
```{r}
hist(as.numeric(Perthtemp))
```
```{r}
library(tidyverse)
#define where you want the breaks in the historgram
userbreak<-c(8,10,12,14,16,18,20,22,24,26)

# remove the ID and site columns
Perthtemp <- Aucitytemp2 %>%
  filter(site=="Perth")

t<-Perthtemp %>%
 dplyr::select(Jan:Dec)

  hist((as.numeric(t)), 
     breaks=userbreak, 
     col="red", 
     main="Histogram of Perth Temperature", 
     xlab="Temperature", 
     ylab="Frequency")
```
```{r}
#Check out the histogram information R generated
histinfo <- as.numeric(t) %>%
  as.numeric()%>%
  hist(.)

histinfo
```
breaks — the cut off points for the bins (or bars), we just specified these
counts — the number of cells in each bin
midpoints — the middle value for each bin
density — the density of data per bin

Check the layer by plotting the geometr
```{r}
plot(Ausoutline$geom)
```
```{r}
#simplify the .shp with lots of points
AusoutSIMPLE <- Ausoutline %>%
  st_simplify(., dTolerance = 1000) %>% #controls the level of generalisation in the units of the map
  st_geometry()%>%
  plot()
```

make sure that both of our layers are in the same coordinate reference system before combine
```{r}
print(Ausoutline)

#this works nicely for rasters
crs(worldclimtemp)
```

```{r}
Austemp <- Ausoutline %>%
  # now crop our temp data to the extent
  terra::crop(worldclimtemp,.)

# plot the output
plot(Austemp)
```
```{r}
exactAus<-terra::mask(Austemp, Ausoutline)
exactAus
```
```{r}
#subset using the known location of the raster
hist(exactAus[[3]], col="red", main ="March temperature")
```
```{r}
#make our raster into a data.frame to be compatible with ggplot2, using a dataframe or tibble
exactAusdf <- exactAus %>%
  as.data.frame()
```

```{r}
library(ggplot2)
# set up the basic histogram
gghist <- ggplot(exactAusdf, 
                 aes(x=Mar)) + 
  geom_histogram(color="black", 
                 fill="white")+
  labs(title="Ggplot2 histogram of Australian March temperatures", 
       x="Temperature", 
       y="Frequency")
# add a vertical line to the hisogram showing mean tempearture
gghist + geom_vline(aes(xintercept=mean(Mar, 
                                        na.rm=TRUE)),
            color="blue", 
            linetype="dashed", 
            size=1)+
  theme(plot.title = element_text(hjust = 0.5))
```
put our variable (months) into a one column using pivot_longer()
```{r}
squishdata<-exactAusdf%>%
  pivot_longer(
  cols = 1:12,
  names_to = "Month",
  values_to = "Temp"
)
```

select two month
```{r}
twomonths <- squishdata %>%
  # | = OR
  filter(., Month=="Jan" | Month=="Jun")
```

```{r}
meantwomonths <- twomonths %>%
  group_by(Month) %>%
  summarise(mean=mean(Temp, na.rm=TRUE))

meantwomonths
```
```{r}
ggplot(twomonths, aes(x=Temp, color=Month, fill=Month)) +
  geom_histogram(position="identity", alpha=0.5)+
  geom_vline(data=meantwomonths, 
             aes(xintercept=mean, 
                 color=Month),
             linetype="dashed")+
  labs(title="Ggplot2 histogram of Australian Jan and Jun
       temperatures",
       x="Temperature",
       y="Frequency")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5))
```
```{r}
data_complete_cases <- squishdata %>%
  drop_na()%>% #dropped all the NAs
  # Month column map in descending order (e.g. Jan, Feb, March..)
  mutate(Month = factor(Month, levels = c("Jan","Feb","Mar",
                                          "Apr","May","Jun",
                                          "Jul","Aug","Sep",
                                          "Oct","Nov","Dec")))

# Plot faceted histogram
ggplot(data_complete_cases, aes(x=Temp, na.rm=TRUE))+
  geom_histogram(color="black", binwidth = 5)+
  labs(title="Ggplot2 faceted histogram of Australian temperatures", 
       x="Temperature",
       y="Frequency")+
  facet_grid(Month ~ .)+
  theme(plot.title = element_text(hjust = 0.5))
```
```{r}
library(plotly)
# split the data for plotly based on month

jan <- squishdata %>%
  drop_na() %>%
  filter(., Month=="Jan")

jun <- squishdata %>%
  drop_na() %>%
  filter(., Month=="Jun")

# give axis titles
x <- list (title = "Temperature")
y <- list (title = "Frequency")

# set the bin width
xbinsno<-list(start=0, end=40, size = 2.5)

# plot the histogram calling all the variables we just set
ihist<-plot_ly(alpha = 0.6) %>%
        add_histogram(x = jan$Temp,
        xbins=xbinsno, name="January") %>%
        add_histogram(x = jun$Temp,
        xbins=xbinsno, name="June") %>% 
        layout(barmode = "overlay", xaxis=x, yaxis=y)

ihist
```

```{r}
# mean per month
meanofall <- squishdata %>%
  group_by(Month) %>%
  summarise(mean = mean(Temp, na.rm=TRUE))

# print the top 1
head(meanofall, n=1)
```

```{r}
# standard deviation per month
sdofall <- squishdata %>%
  group_by(Month) %>%
  summarize(sd = sd(Temp, na.rm=TRUE))

# maximum per month
maxofall <- squishdata %>%
  group_by(Month) %>%
  summarize(max = max(Temp, na.rm=TRUE))

# minimum per month
minofall <- squishdata %>%
  group_by(Month) %>%
  summarize(min = min(Temp, na.rm=TRUE))

# Interquartlie range per month
IQRofall <- squishdata %>%
  group_by(Month) %>%
  summarize(IQR = IQR(Temp, na.rm=TRUE))

# perhaps you want to store multiple outputs in one list..
lotsofstats <- squishdata %>%
  group_by(Month) %>%
  summarize(IQR = IQR(Temp, na.rm=TRUE), 
            max=max(Temp, na.rm=T))

# or you want to know the mean (or some other stat) 
#for the whole year as opposed to each month...

meanwholeyear=squishdata %>%
  summarize(meanyear = mean(Temp, na.rm=TRUE))
```


Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
